Stability and accuracy of free surface time integration in viscous flows

### 2016-08-03 free-surface-paper.tex
Fix up references.

%% main text
\section{Introduction}
\label{sec:intro}

@@

The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
which should be zero when evaluated on a true free surface.
Rather than analyzing finite deformation to the free surface (a nonlinear problem),
we will make the analytically useful approximation of small deformations about the hydrostatic 
reference surface and analyze their stability.
We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
over the hydrostatic reference surface and introduce a temporary auxiliary variable $\zeta$ which 
represents the (small) topography of the free surface relative to that reference surface.
On the reference surface the gravity vector is opposite the direction of the surface normal $\mathbf{g} = -g \mathbf{n}$.
The stress on this reference surface can be approximated by using the first order Taylor series
expansion of the hydrostatic pressure profile:
\begin{equation}
\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = -\rho_0 g \zeta \mathbf{I}.
\label{eq:hydrostatic}
\end{equation}
Equation~\eqref{eq:weak_stokes} then becomes

@@

\section{The relaxation spectrum is positive definite}
\label{app:positive_definite}
On physical grounds, we expect the relaxation spectrum described in Section~\ref{sec:eigenvalue}
to have only positive eigenvalues $\tau_i$.
An imaginary eigenvalue would have an oscillitory component,
which we do not expect in creeping flows (oscillation being usually due to inertia). 
A negative eigenvalue would grow exponentially, which would violate conservation of energy.
Here we demonstrate the positivity of the relaxation time spectrum.
We could proceed with either the continuous or the discretized version of the governing
equations, but choose the continuous form here.
We begin with the homogeneous Stokes equation (Equation~\eqref{eq:stokes}):
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}.
\label{eq:homogeneous_stokes}
\end{equation}
We can transform this into the mechanical energy equation by multiplying it by $\mathbf{u}$
and integrating over the domain:
\begin{equation}
\int_{\Omega(t)} \mathbf{u} \cdot \nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \int_{\Omega(t)} \mathbf{u}\cdot \nabla p = 0.
\end{equation}
Integrating by parts and applying boundary conditions yields
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Omega(t)}p \nabla \cdot \mathbf{u} - \int_{\Gamma_F(t)} \mathbf{u} \cdot \mathbf{T} \cdot \mathbf{n} = 0.
\end{equation}
Where the integrals correspond to the viscous dissipation, the work done by pressure, and the work done by boundary stresses, respectively.
Since we are using the incompressible approximation, the second integral is zero.
On a true free surface the boundary stresses are zero, but we make the same approximation
used in Equation~\eqref{eq:hydrostatic} and map the stress onto the reference surface, finding
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Gamma_F(t)} \rho_0 g \zeta \mathbf{u} \cdot \mathbf{n} = 0.
\end{equation}
We can now eliminate $\zeta$ from the equation in the same way that was used in Equation~\eqref{eq:weak_eigen},
thereby introducing the relaxation time $\tau_i$:
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2 = 0.
\end{equation}
This equation may be solved for $\tau_i$:
\begin{equation}
\tau_i = \frac{\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})}{ \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2}.
\end{equation}
The top and bottom of the right hand side are both strictly positive numbers so the $\tau_i$ must also be positive,
demonstrating the positive definiteness of the system. A similar analysis can be performed for the discrete system.
There is the possibility for velocities which do not move the free surface at all, in which case the bottom
integral goes to zero, and the relaxation time becomes infinite. One example of this is the case of shear flow with
a free upper surface and periodic horizontal boundary conditions. Since we are usually interested in the smallest
relaxation times, these infinite eigenvalues are not usually a problem.

>>>>

%% main text
\section{Introduction}
\label{sec:introduce}

@@

The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
which should be zero when evaluated on a true free surface.
Rather than analyzing finite deformation to the free surface (a nonlinear problem),
we will make the analytically useful approximation of small deformations about the hydrostatic 
reference surface and analyze their stability.
We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
over the hydrostatic reference surface and introduce a temporary auxiliary variable $\zeta$ which 
represents the (small) topography of the free surface relative to that reference surface.
On the reference surface the gravity vector is opposite the direction of the surface normal $\mathbf{g} = -g \mathbf{n}$.
The stress on this reference surface can be approximated by using the first order Taylor series
expansion of the hydrostatic pressure profile:
\begin{equation}
\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = -\rho_0 g \zeta \mathbf{I}.
\label{eq:hydrostat}
\end{equation}
Equation~\eqref{eq:weak_stokes} then becomes

@@

\section{The relaxation spectrum is positive definite}
\label{app:positive_definite}
On physical grounds, we expect the relaxation spectrum described in Section~\ref{sec:eigenvalue}
to have only positive eigenvalues $\tau_i$.
An imaginary eigenvalue would have an oscillitory component,
which we do not expect in creeping flows (oscillation being usually due to inertia). 
A negative eigenvalue would grow exponentially, which would violate conservation of energy.
Here we demonstrate the positivity of the relaxation time spectrum.
We could proceed with either the continuous or the discretized version of the governing
equations, but choose the continuous form here.
We begin with the homogeneous Stokes equation (Equation~\eqref{eq:stokes}):
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}.
\end{equation}
We can transform this into the mechanical energy equation by multiplying it by $\mathbf{u}$
and integrating over the domain:
\begin{equation}
\int_{\Omega(t)} \mathbf{u} \cdot \nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \int_{\Omega(t)} \mathbf{u}\cdot \nabla p = 0.
\end{equation}
Integrating by parts and applying boundary conditions yields
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Omega(t)}p \nabla \cdot \mathbf{u} - \int_{\Gamma_F(t)} \mathbf{u} \cdot \mathbf{T} \cdot \mathbf{n} = 0.
\end{equation}
Where the integrals correspond to the viscous dissipation, the work done by pressure, and the work done by boundary stresses, respectively.
Since we are using the incompressible approximation, the second integral is zero.
On a true free surface the boundary stresses are zero, but we make the same approximation
used in Equation~\eqref{eq:hydrostat} and map the stress onto the reference surface, finding
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Gamma_F(t)} \rho_0 g \zeta \mathbf{u} \cdot \mathbf{n} = 0.
\end{equation}
We can now eliminate $\zeta$ from the equation in the same way that was used in Equation~\eqref{eq:weak_eigen},
thereby introducing the relaxation time $\tau_i$:
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2 = 0.
\end{equation}
This equation may be solved for $\tau_i$:
\begin{equation}
\tau_i = \frac{\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})}{ \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2}.
\end{equation}
The top and bottom of the right hand side are both strictly positive numbers so the $\tau_i$ must also be positive,
demonstrating the positive definiteness of the system. A similar analysis can be performed for the discrete system.
There is the possibility for velocities which do not move the free surface at all, in which case the bottom
integral goes to zero, and the relaxation time becomes infinite. One example of this is the case of shear flow with
a free upper surface and periodic horizontal boundary conditions. Since we are usually interested in the smallest
relaxation times, these infinite eigenvalues are not usually a problem.
